---
title: 'Bios 6301: Assignment 2'
author: "Ying Ji"
output: pdf_document
---

**Grade: 55/50**

*(informally) Due Tuesday, 20 September, 1:00 PM*


50 points total.

This assignment won't be submitted until we've covered Rmarkdown.
Create R chunks for each question and insert your R code appropriately.
Check your output by using the `Knit PDF` button in RStudio.

1. **Working with data** In the `datasets` folder on the course GitHub repo, you will find a file called `cancer.csv`, which is a dataset in comma-separated values (csv) format. This is a large cancer incidence dataset that summarizes the incidence of different cancers for various subgroups. (18 points)

    1. Load the data set into R and make it a data frame called `cancer.df`. (2 points)
```{r}
#setwd("~/Documents/biostat_computing/bios6301_homework/")
cancer.df<-data.frame(read.csv("cancer.csv"))
```


    2. Determine the number of rows and columns in the data frame. (2)
```{r}
dim(cancer.df)
```


3. Extract the names of the columns in `cancer.df`. (2)
```{r}
colnames(cancer.df)
```

    4. Report the value of the 3000th row in column 6. (2)
```{r}
cancer.df[3000,6]
```



    5. Report the contents of the 172nd row. (2)
```{r}
cancer.df[172,]
```

    6. Create a new column that is the incidence *rate* (per 100,000) for each row.(3)
```{r}
cancer.df[,"incidence_rate"]<-cancer.df[,"incidence"]/cancer.df[,"population"]*100000
```

    7. How many subgroups (rows) have a zero incidence rate? (2)
```{r}
sum(cancer.df[,"incidence_rate"]==0)
```

    8. Find the subgroup with the highest incidence rate.(3)
```{r}
which.max(cancer.df[,"incidence_rate"])
```



2. **Data types** (10 points)

1. Create the following vector: `x <- c("5","12","7")`. Which of the following commands will produce an error message? For each command, Either explain why they should be errors, or explain the non-erroneous result. (4 points)
    
```
max(x):returned "7"", not the max number 12. The reason is that x is a character vector, so the order depends on the first number of each element: 7>5>1
sort(x): returned ("12,"5,"7"), sort return the order of the vector, from first number of each element:7>5>1
sum(x): produced an error "invalid type of arugment", because the argument should be numeric type
```
2. For the next two commands, either explain their results, or why they should produce errors. (3 points)
```
y <- c("5",7,12)
y[2] + y[3]
produced errors: non-numeric argument to binary operator
y is a vector with character "5" and numeric element 7 and 12, after combination using c() the type of vector coerced to the least flexible form: character, then y[2] and y[3] are all characters, cannot by applied "+"
```

3. For the next two commands, either explain their results, or why they should produce errors. (3 points)

z <- data.frame(z1="5",z2=7,z3=12)
z[1,2] + z[1,3]
```
result is 19, because z[1,2] and z[1,3] are all of numeric type. Data.frame keep the original class of each element, not like c(). So can be applied "+"
```

3. **Data structures** Give R expressions that return the following matrices and vectors (*i.e.* do not construct them manually). (3 points each, 12 total)

    1. $(1,2,3,4,5,6,7,8,7,6,5,4,3,2,1)$
```{r}
c(seq(8),seq(7,1)) 
```


    2. $(1,2,2,3,3,3,4,4,4,4,5,5,5,5,5)$
```{r}
rep(1:5,times=1:5)
```


    3. $\begin{pmatrix}
      0 & 1 & 1  \\
      1 & 0 & 1  \\
      1 & 1 & 0  \\
    \end{pmatrix}$
```{r}
pmatrix<-matrix(1,nrow=3,ncol=3)
diag(pmatrix)<-0
pmatrix
```


    4. $\begin{pmatrix}
      1 & 2 & 3 & 4 \\
      1 & 4 & 9 & 16 \\
      1 & 8 & 27 & 64  \\
      1 & 16 & 81 & 256 \\
      1 & 32 & 243 & 1024  \\
    \end{pmatrix}$
```{r}
a<-seq(4)
pmatrix<-rbind(a,a^2,a^3,a^4,a^5)
pmatrix
```


4. **Basic programming** (10 points)

    1. Let $h(x,n)=1+x+x^2+\ldots+x^n = \sum_{i=0}^n x^i$. Write an R program to calculate $h(x,n)$ using a `for` loop. (5 points)
```{r}
h <- function(x,n) {
  s <- 0
  for (i in 0:n) {
    s= s + x^i
  }
  return(s)
}
```



    1. If we list all the natural numbers below 10 that are multiples of 3 or 5, we get 3, 5, 6 and 9. The sum of these multiples is 23. Write an R program to perform the following calculations. (5 points)

        1. Find the sum of all the multiples of 3 or 5 below 1,000. (3, [euler1])
```{r}
s<-0
for (i in 1:1000-1){
        if (i %%3 ==0|i %% 5==0){
        
        s=s+i
        
        }
}
s
```


1. Find the sum of all the multiples of 4 or 7 below 1,000,000. (2)
```{r}
s<-0

for (i in 1:1000000-1){
        if (i %%4 ==0|i %% 7==0){
        
        s=s+i
        }
}
s
```


    1. Each new term in the Fibonacci sequence is generated by adding the previous two terms. By starting with 1 and 2, the first 10 terms will be $(1, 2, 3, 5, 8, 13, 21, 34, 55, 89)$. Write an R program to calculate the sum of the first 15 even-valued terms. (5 bonus points, [euler2])

```{r}
f<-numeric(50)
j<-1
f[1]<-1
f[2]<-2
s<-f[1]+f[2]
for (i in 3:50){
        f[i]<-f[i-2]+f[i-1]
        if (f[i] %%2 ==0 && j<15){
        s=s+f[i]
        j=j+1
        }
}
s
```

**JC Grading +5 Bonus**

Some problems taken or inspired by projecteuler.

[euler1]:https://projecteuler.net/problem=1
[euler2]: https://projecteuler.net/problem=2
